---
title: "Aboriginal Employment in Victoria"
author: "Strategy and Priority Project - Data and Analytics"
format: 
  html:
    embed-resources: true
editor: visual
date: "`r Sys.Date()`"
toc: true
copyright: 
  text: "Victorian Government Department of Jobs, Skills, Industry and Regions"
# license: CC BY
execute:
  echo: false
  warning: false
abstract: 
  "The following data estimates Aboriginal employment statistics for Victorian regions as of `r format(readabs::check_latest_date(cat_no = '6291.0.55.001', tables = 16), '%B %Y')`. This data reflects [2021 census employment](https://www.abs.gov.au/statistics/microdata-tablebuilder/tablebuilder) levels for Aboriginal peoples, adjusted for [fluctuations in regional employment levels](https://www.abs.gov.au/statistics/labour/employment-and-unemployment/labour-force-australia-detailed) that have occurred since August 2021. As the Bureau of Statistics does not report  regional statistics for Aboriginal people over time, any changes in employment since August 2021 only reflect the broader region's employment trend, not improvement or deterioration in employment outcomes specific to Aboriginal peoples."
---

```{r output=FALSE}
# Packages
library(dplyr)
library(tidyr)
library(readr)
library(readabs)
library(leaflet)
library(sf)
library(strayr)
library(stringr)
library(slider)
library(htmltools)
library(highcharter)
library(writexl)
library(DT)

# Get data
census <- read_csv("data/census.csv")
lfs_raw <- read_abs(cat_no = "6291.0.55.001", tables = 16) %>% separate_series()
sa3 <- read_absmap(area = "sa3", year = 2021)

# Find vic SA3
vic_sa4 <- sa3 %>% 
  filter(state_name_2021 == "Victoria") %>% 
  pull(sa4_name_2021) %>% 
  unique() %>% 
  str_subset("No usual address|Migratory", negate = T) %>% 
  str_replace("^North West$", "Victoria - North West")

# clean LFS
lfs <- lfs_raw %>% 
  filter(series_1 %in% vic_sa4, series_3 == "Persons") %>% 
  select(date, sa4 = series_1, measure = series_2, value) %>% 
  pivot_wider(names_from = measure, values_from = value) %>%
  group_by(sa4) %>% 
  arrange(date) %>% 
  mutate(
    across(
      all_of(
        c(
          "Unemployment rate", 
          "Participation rate", 
          "Employment to population ratio"
          )
        ), 
      ~slide_mean(., before = 2)
      )
    ) %>% 
  filter(date >= "2020-08-01") %>% 
  transmute(
    date,
    sa4 = recode(sa4, "Victoria - North West" = "North West"),
    unemp_rate_grow = `Unemployment rate` / first(`Unemployment rate`),
    part_rate_grow = `Participation rate` / first(`Participation rate`),
    emp_pop_grow = `Employment to population ratio` / 
      first(`Employment to population ratio`)
  ) %>% 
  ungroup() #%>% 
  # mutate(across(contains("grow"), ~ . / 100))

# combine data and scale over time
census <- mutate(census, sa3 = as.character(sa3))
combo <- right_join(sa3, census, by = c("sa3_code_2021" = "sa3")) %>% 
  full_join(
    lfs, 
    by = c("sa4_name_2021" = "sa4"), 
    relationship = "many-to-many"
    ) %>% 
  mutate(
    unemp_rate = unemp_rate * unemp_rate_grow,
    part_rate = part_rate * part_rate_grow,
    emp_pop = emp_pop * emp_pop_grow
  )
```

## Headline employment statistics
The following charts show estimate Victoria-wide Aboriginal employment statistics and compare them with broader population values. Please note, census-based employment statistics will not reconcile with official labour statistics. For more information [please read the data quality statement](#data-quality). 

```{r}

lfs_scalars <- lfs_raw %>%
  filter(date >= "2020-08-01", series_3 == "Persons", series_1 == "Victoria") %>%
  select(date, measure = series_2, value) %>%
  pivot_wider(names_from = measure, values_from = value) %>%
  arrange(date) %>%
  transmute(
    date,
    unemp_rate_grow = `Unemployment rate` / first(`Unemployment rate`),
    part_rate_grow = `Participation rate` / first(`Participation rate`),
    emp_pop_grow = `Employment to population ratio` /
      first(`Employment to population ratio`)
  ) %>%
  filter(date == max(date))

headline_data <- census %>%
  group_by(ind) %>%
  summarise(
    employed = sum(employed),
    unemployed = sum(unemployed),
    nilf = sum(nilf)
  ) %>%
  ungroup() %>%
  mutate(
    unemp_rate = unemployed / (employed + unemployed),
    part_rate = (employed + unemployed) / (employed + unemployed + nilf),
    emp_pop = employed / (employed + unemployed + nilf),
    unemp_rate = unemp_rate * lfs_scalars$unemp_rate_grow,
    part_rate = part_rate * lfs_scalars$part_rate_grow,
    emp_pop = emp_pop * lfs_scalars$emp_pop_grow
  ) %>% 
  filter(ind != "Non-Aboriginal") %>% 
  mutate(ind = recode(ind, total = "All Vic"))

hc_unemp <- headline_data %>% 
  hchart(
    "column", 
    hcaes(ind, unemp_rate, name = ind), 
    name = "Unemployment rate"
    ) %>% 
  hc_title(
    text = "Unemployment rate"
  ) %>% 
  hc_tooltip(
    formatter = JS(
      "function () {
            return '<strong>' + this.point.name + '</strong></br>'+
            'Unemployment rate: ' + Math.round(this.y * 1000) / 10 + '%';
        }"
    )
  ) %>% 
  hc_yAxis(
    title = list(text = NULL, enabled = FALSE),
    labels = list(
      formatter = JS(
        "function () {
            return Math.round(this.value * 1000) / 10 + '%';
        }"
      )
         
    )
  ) %>% 
  hc_xAxis(
    title = list(text = NULL)
  )

hc_part <- headline_data %>% 
  hchart(
    "column", 
    hcaes(ind, part_rate, name = ind), 
    name = "Participation rate"
    ) %>% 
  hc_title(
    text = "Participation rate"
  ) %>% 
  hc_tooltip(
    formatter = JS(
      "function () {
            return '<strong>' + this.point.name + '</strong></br>'+
            'Participation rate: ' + Math.round(this.y * 1000) / 10 + '%';
        }"
    )
  ) %>% 
  hc_yAxis(
    title = list(text = NULL, enabled = FALSE),
    labels = list(
      formatter = JS(
        "function () {
            return Math.round(this.value * 100) + '%';
        }"
      )
         
    )
  ) %>% 
  hc_xAxis(
    title = list(text = NULL)
  )

hc_emp_pop <- headline_data %>% 
  hchart(
    "column", 
    hcaes(ind, emp_pop, name = ind), 
    name = "Employment to population ratio"
    ) %>% 
  hc_title(
    text = "Employment to population ratio"
  ) %>% 
  hc_tooltip(
    formatter = JS(
      "function () {
            return '<strong>' + this.point.name + '</strong></br>'+
            'Employment to population ratio: ' + Math.round(this.y * 1000) / 10 + '%';
        }"
    )
  ) %>% 
  hc_yAxis(
    title = list(text = NULL, enabled = FALSE),
    labels = list(
      formatter = JS(
        "function () {
            return Math.round(this.value * 100) + '%';
        }"
      )
         
    )
  ) %>% 
  hc_xAxis(
    title = list(text = NULL)
  )

# hw_grid(hc_unemp, hc_part, hc_emp_pop, ncol = 3, add_htmlgrid_css = T)
hc_unemp
hc_part
hc_emp_pop
```

## Maping employment

The following provides unemployment rates, participation rates and estimate working-age population of Aboriginal peoples by region. Regions are defined using the [Australian Bureau of Statistics Statistical Area level 3](https://www.abs.gov.au/statistics/statistical-geography/australian-statistical-geography-standard-asgs). Please note that due to sample size issues, some regions may report an unemployment rate of zero. These observations likely represent statistical anomalies rather than underlying conditions.

```{r}
mapdata <- combo %>% 
  group_by(sa3_name_2021, ind) %>% 
  mutate(
    unemp_rate_flag = unemp_rate < 0.01 | unemp_rate > 1,
    unemp_rate = ifelse(unemp_rate_flag, first(unemp_rate), unemp_rate),
    part_rate_flag = part_rate < 0.2 | part_rate > 1,
    part_rate = ifelse(part_rate_flag, first(part_rate), part_rate),
  ) %>% 
  ungroup() %>% 
  filter(date == max(date, na.rm = T), ind == "Aboriginal") %>% 
  mutate(
    label = sprintf(
        "<strong style=\"font-size:1rem;\">%s</strong>
        <p>Estimate Aboriginal labour statistics:</p>
        <ul style = \"padding-left: 1.2em;\">
          <li>Unemployment rate: %s%s</li>
          <li>Participation rate: %s%s</li>
          <li>Census population >15yo: %s</li>
        </ul>
        %s",
        sa3_name_2021,
        paste0(round(unemp_rate, 3) * 100, "%"),
        ifelse(unemp_rate_flag, "*", ""),
        paste0(round(part_rate, 3) * 100, "%"),
        ifelse(part_rate_flag, "*", ""),
        scales::comma(round(nilf + unemployed + employed)),
        ifelse(
          unemp_rate_flag|part_rate_flag, 
          "<p><small>* 2021 census value used as forecast non-viable</small><p>",
          ""
          )
      )
  )

pal <- colorNumeric("viridis", domain = mapdata$unemp_rate)

mapdata %>% 
  leaflet() %>% 
  addProviderTiles("Esri.WorldGrayCanvas") %>% 
  addPolygons(
    fillColor = ~pal(unemp_rate),
    fillOpacity = 0.8,
    weight = 1,
    color = "#FFFFFF",
    label = ~lapply(label, HTML),
    highlightOptions = highlightOptions(
      weight = 3, 
      opacity = 1
    )
  ) %>% 
  addLegend(
    pal = pal,
    values = ~unemp_rate, 
    title = "Unemployment</br>Rate",
    labFormat = labelFormat(suffix = "%", transform = \(x) round(x * 100, 1))
  )
```

The following chart demonstrates how unemployment rates vary with distance from Melbourne.

```{r}
melb <- mapdata %>% 
  filter(sa3_name_2021 == "Melbourne City") %>% 
  st_centroid()

chart_data <- mapdata %>% 
  st_centroid() %>% 
  mutate(
    distance = st_distance(geometry, melb),
    distance = as.numeric(distance) / 1000
    )

trend_unemp <- loess(unemp_rate ~ distance, data = chart_data, span = 0.75)
trend_unemp <- data.frame(
  x = as.vector(trend_unemp$x), 
  y = trend_unemp$fitted
  ) %>% 
  arrange(x) %>% 
  highcharter::list_parse()

hchart(
  chart_data, 
  type = "point", 
  hcaes(x = distance, y = unemp_rate, name = sa3_name_2021),
  name = "Regional unemployment rate"
  ) %>% 
  hc_add_series(
    data = trend_unemp, 
    type = "spline", 
    name = "Trend", 
    enableMouseTracking = FALSE
    ) %>% 
  hc_title(
    text = "Unemployment rates are higher in Victoria's regional areas"
  ) %>% 
  hc_subtitle(
    text = "Regional Aboriginal unemployment rates by distance from Melbourne and smoothed trend line"
  ) %>% 
  hc_tooltip(
    formatter = JS(
      "function () {
            return '<strong>' + this.point.name + '</strong></br>'+
            'Unemployment rate: ' + Math.round(this.y * 1000) / 10 + '%</br>' +
            'Distance from Melbourne: ' + Math.round(this.x) + 'km';
        }"
    )
  ) %>% 
  hc_yAxis(
    title = list(text = "Unemployment rate"),
    labels = list(
      formatter = JS(
        "function () {
            return Math.round(this.value * 1000) / 10 + '%';
        }"
      )
         
    )
  ) %>% 
  hc_xAxis(
    title = list(text = "Distance from Melbourne"),
    labels = list(
      formatter = JS(
        "function () {
            return Math.round(this.value) + 'km';
        }"
      )
    )
  )

```

The following chart shows a similar trend with participation rates.

```{r}
trend_part <- loess(part_rate ~ distance, data = chart_data, span = 0.75)
trend_part <- data.frame(
  x = as.vector(trend_part$x), 
  y = trend_part$fitted
  ) %>% 
  arrange(x) %>% 
  highcharter::list_parse()

hchart(
  chart_data, 
  type = "point", 
  hcaes(x = distance, y = part_rate, name = sa3_name_2021),
  name = "Regional participation rate"
  ) %>% 
  hc_add_series(
    data = trend_part, 
    type = "spline", 
    name = "Trend", 
    enableMouseTracking = FALSE
    ) %>% 
  hc_title(
    text = "Participation rates are lower in Victoria's regional areas"
  ) %>% 
  hc_subtitle(
    text = "Regional Aboriginal participation rates by distance from Melbourne and smoothed trend line"
  ) %>% 
  hc_tooltip(
    formatter = JS(
      "function () {
            return '<strong>' + this.point.name + '</strong></br>'+
            'Participation rate: ' + Math.round(this.y * 100) + '%</br>' +
            'Distance from Melbourne: ' + Math.round(this.x) + 'km';
        }"
    )
  ) %>% 
  hc_yAxis(
    title = list(text = "Participation rate"),
    labels = list(
      formatter = JS(
        "function () {
            return Math.round(this.value * 1000) / 10 + '%';
        }"
      )
         
    )
  ) %>% 
  hc_xAxis(
    title = list(text = "Distance from Melbourne"),
    labels = list(
      formatter = JS(
        "function () {
            return Math.round(this.value) + 'km';
        }"
      )
    )
  )
```

## Data downloads

[Full data is available for download here](https://github.com/djsir-data/aboriginallabourforce/raw/main/data/Vic%20Aboriginal%20employment.xlsx). Alternatively, you can manually explore the data in the following table.

```{r}
out_data <- combo %>% 
  st_drop_geometry() %>% 
  filter(ind == "Aboriginal") %>% 
  arrange(desc(date), sa3_name_2021) %>% 
  select(
    `Date` = date,
    `Region` = sa3_name_2021,
    `Unemployment rate` = unemp_rate,
    `Participation rate` = part_rate,
    `Employment to population ratio` = emp_pop
  )

write_xlsx(out_data, "data/Vic Aboriginal employment.xlsx")

out_data %>% 
  mutate(across(where(is.numeric), ~ifelse(. == 0 | . > 1, as.numeric(NA), .))) %>% 
  mutate(across(where(is.numeric), ~round(., 3) * 100)) %>% 
  mutate(
    across(where(is.numeric), ~ifelse(is.na(.), ., paste0(., "%"))),
    Date = format(Date, "%b %Y")
    ) %>% 
  datatable(rownames = F)
```

## Data quality statement {#data-quality}

The information contained in this report attempts to bridge two knowledge gaps; Aboriginal employment rates since census and fine geographical disaggregation of Aboriginal employment. These new features may introduce the following errors:

-   **Small regions with few Aboriginal people may report anomalous results**\
    For example, if a given region has few Aboriginal people but they are all employed, the region would report an unemployment rate of zero. Though technically correct for the snapshot of the region during census, this result does not represent the underlying conditions or likelihood of employment for Aboriginal people living here.

-   **Movements in employment since Census don't capture Aboriginal-specific changes**\
    While census provides a reasonably accurate picture of Aboriginal employment conditions, the [Australian Bureau of Statistics time series employment publication](https://www.abs.gov.au/statistics/labour/employment-and-unemployment/labour-force-australia-detailed) does not provide any detailed Aboriginal information. This means that if Aboriginal employment conditions improved or deteriorated at a different rate to the general public, this report will not reflect the additional changes.

-   **Census employment rates do not reconcile with official employment statistics**\
    As [noted by the Australian Bureau of Statistics](https://www.abs.gov.au/statistics/detailed-methodology-information/information-papers/comparing-2021-census-and-labour-force-survey), Census-based employment statistics will not reconcile with Victoria's official employment statistics. This is not to say Census statistics are inferior; it simply has a different sampling methodology.

Considering the above points, *the data contained within this report should be used for indicative purposes only*. No claim is made as to the accuracy or currency of any of the content in this report at any time. The Victorian Government and this agency (the Department of Jobs, Skills, Industry and Regions) do not accept any liability to any person for the information (or the use of such information) which is provided in this report, including information sourced or derived from third parties or incorporated into the repot by reference.
