---
title: "ATSI Employment in Victoria"
author: "Strategy and Priority Project - Data and Analytics"
format: html
editor: visual
date: "`r Sys.Date()`"
execute:
  echo: false
abstract: 
  "The following data estimates Aboriginal and Torres Straight Islander (ATSI) unemployment rates for Victorian regions. This data reflects [2021 census employment](https://www.abs.gov.au/statistics/microdata-tablebuilder/tablebuilder) levels for ATSI peoples, adjusted for [fluctuations in regional employment levels](https://www.abs.gov.au/statistics/labour/employment-and-unemployment/labour-force-australia-detailed) that have occurred since August 2021. As the Bureau of Statistics does not report regional statistics for ATSI people at a regional level over time, any changes in employment since August 2021 only reflect the broader region's employment trend, not improvement or deterioration in employment outcomes specific to ATSI people."
---

```{r output=FALSE}
# Packages
library(dplyr)
library(tidyr)
library(readr)
library(readabs)
library(leaflet)
library(sf)
library(strayr)
library(stringr)
library(slider)
library(htmltools)

# Get data
census <- read_csv("data/census.csv")
lfs <- read_abs(cat_no = "6291.0.55.001", tables = 16)
sa3 <- read_absmap(area = "sa3", year = 2021)

# Find vic SA3
vic_sa4 <- sa3 %>% 
  filter(state_name_2021 == "Victoria") %>% 
  pull(sa4_name_2021) %>% 
  unique() %>% 
  str_subset("No usual address|Migratory", negate = T) %>% 
  str_replace("^North West$", "Victoria - North West")

# clean LFS
lfs <- lfs %>% 
  separate_series() %>% 
  filter(date >= "2020-08-01", series_3 == "Persons", series_1 %in% vic_sa4) %>% 
  select(date, sa4 = series_1, measure = series_2, value) %>% 
  pivot_wider(names_from = measure, values_from = value) %>% 
  transmute(
    date,
    sa4 = recode(sa4, "Victoria - North West" = "North West"),
    employed_grow = `Employed total`,
    unemployed_grow = `Unemployed total`,
    nilf_grow = 
      `Labour force total` / `Participation rate` - `Labour force total`
  ) %>% 
  group_by(sa4) %>% 
  arrange(date) %>% 
  mutate(across(contains("grow"), ~ . / first(.))) %>% 
  mutate(across(contains("grow"), ~ slide_mean(., before = 2))) %>% 
  ungroup()

# combine data and scale over time
census <- mutate(census, sa3 = as.character(sa3))
combo <- right_join(sa3, census, by = c("sa3_code_2021" = "sa3")) %>% 
  full_join(
    lfs, 
    by = c("sa4_name_2021" = "sa4"), 
    relationship = "many-to-many"
    ) %>% 
  mutate(
    employed = employed * employed_grow,
    unemployed = unemployed * unemployed_grow,
    nilf_grow = nilf * nilf_grow,
    unemp_rate = unemployed / (employed + unemployed),
    part_rate = (employed + unemployed) / (employed + unemployed + nilf),
    emp_pop = employed / (employed + unemployed + nilf)
  )
```

## Maping employment
The following provides unemployment rates, participation rates and estimate working-age population of ATSI peoples by region. Regions are defined using the [Australian Bureau of Statistics Statistical Area level 3](https://www.abs.gov.au/statistics/statistical-geography/australian-statistical-geography-standard-asgs). Please note that due to sample size issues, some regions may report an unemployment rate of zero. These observations likely represent statistical anomalies rather than underlying conditions. 
```{r}
mapdata <- combo %>% 
  filter(date == max(date, na.rm = T), ind == "ATSI") %>% 
  mutate(
    label = sprintf(
        "<strong style=\"font-size:1rem;\">%s</strong>
        <p>Estimate ATSI labour statistics:</p>
        <ul style = \"padding-left: 1.2em;\">
          <li>Unemployment rate: %s</li>
          <li>Participation rate: %s</li>
          <li>Population >15 years old: %s</li>
        </ul>",
        sa3_name_2021,
        paste0(round(unemp_rate, 3) * 100, "%"),
        paste0(round(part_rate, 3) * 100, "%"),
        scales::comma(round(nilf + unemployed + employed))
      )
  )

pal <- colorNumeric("viridis", domain = mapdata$unemp_rate)

mapdata %>% 
  leaflet() %>% 
  addProviderTiles("Esri.WorldGrayCanvas") %>% 
  addPolygons(
    fillColor = ~pal(unemp_rate),
    fillOpacity = 0.8,
    weight = 1,
    color = "#FFFFFF",
    label = ~lapply(label, HTML),
    highlightOptions = highlightOptions(
      weight = 3, 
      opacity = 1
    )
  ) %>% 
  addLegend(
    pal = pal,
    values = ~unemp_rate, 
    title = "Unemployment</br>Rate",
    labFormat = labelFormat(suffix = "%", transform = \(x) round(x * 100, 1))
  )
```

